---
title: "Getting started with survkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with survkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1L)
```

`survkit` gives one interface over the established survival stack. You choose a
method by name; the toolkit picks the backend, runs a small-sample power check,
and hands back a single typed object that prints, summarises, predicts and tidies
the same way regardless of which estimator produced it.

```{r}
library(survkit)
```

## What can it fit?

The method registry is the catalogue. Each row is an estimator; `available`
reports whether its backend is installed.

```{r}
survkit_methods()
```

## A first fit

A Weibull accelerated-failure-time model on the lung-cancer data:

```{r}
fit <- survkit(survival::Surv(time, status) ~ age + sex, survival::lung,
               method = "weibull")
fit
```

`survkit_tidy()` gives the coefficient table with ratios and confidence
intervals; `survkit_curve()` gives predicted survival on a time grid.

```{r}
survkit_tidy(fit)
survkit_curve(fit, times = c(100, 300, 500))
```

## Switching method, same interface

The Cox proportional-hazards fit reads the same and tidies the same:

```{r}
cox <- survkit(survival::Surv(time, status) ~ age + sex, survival::lung,
               method = "cox")
survkit_tidy(cox)
```

## The small-sample gate

Survival models identify from events, not rows. On a small subset the
events-per-variable gate flags the fit as under-powered:

```{r}
small <- survival::lung[1:25, ]
survkit_power(survival::Surv(time, status) ~ age + sex + ph.ecog, small)
```

`survkit()` runs this automatically and warns, so you can simplify, penalise, or
abstain rather than trust over-fit estimates.

## Competing risks

With a multi-state response (the event is a factor whose first level is
censoring), the Fine-Gray subdistribution hazard is one call:

```{r}
mg <- survival::mgus2
mg$etime <- with(mg, ifelse(pstat == 1, ptime, futime))
mg$event <- with(mg, factor(ifelse(pstat == 1, "pcm",
                            ifelse(death == 1, "death", "censor")),
                            levels = c("censor", "pcm", "death")))
fg <- survkit(survival::Surv(etime, event) ~ age + sex, mg,
              method = "fine_gray", cause = "pcm")
survkit_tidy(fg)
```

## Cumulative incidence

For competing risks the Aalen-Johansen cumulative incidence of each event is one
call, and it accounts for the competing event rather than treating it as
censoring:

```{r}
ci <- survkit_cuminc(survival::Surv(etime, event) ~ 1, mg)
head(ci)
```

## Decision-relevant summaries

The restricted mean survival time is the area under the survival curve to a
horizon, a summary that stays defined when the median is never reached:

```{r}
km <- survkit(survival::Surv(time, status) ~ 1, survival::lung, method = "km")
survkit_rmst(km, tau = 365)
```

Predictions can be scored for discrimination and calibration. The concordance,
the censoring-weighted Brier score, and its index of prediction accuracy come
back together:

```{r}
survkit_score(cox, survival::lung, times = c(200, 400, 600))
```

## Triangulating across methods

Fitting the same question several ways is the honesty check survkit is built
for. When the methods disagree on what a decision turns on, at least one of them
is misspecified:

```{r}
survkit_triangulate(survival::Surv(time, status) ~ age + sex, survival::lung,
                    methods = c("weibull", "cox", "km"), tau = 500)
```

## Extending the toolkit

Adding a method is one registry call; the public interface does not change.

```{r}
survkit_register("exp_survreg", kind = "parametric", backend = "survival",
  fit = function(formula, data, ...) {
    raw <- survival::survreg(formula, data = data, dist = "exponential", ...)
    list(raw = raw, coef = coef(raw), vcov = vcov(raw),
         loglik = as.numeric(logLik(raw)), n = nrow(data),
         n_events = NA_integer_, kind = "parametric", backend = "survival")
  })
"exp_survreg" %in% survkit_methods()$method
```
