---
title: "Getting started with survkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with survkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4.2, dpi = 110)
set.seed(1L)
```

`survkit` gives one interface over the established survival stack. You choose a
method by name; the toolkit picks the backend, runs a small-sample power check,
and hands back a single typed object that prints, summarises, predicts and tidies
the same way regardless of which estimator produced it.

```{r}
library(survkit)
```

## What can it fit?

The method registry is the catalogue. Each row is an estimator; `available`
reports whether its backend is installed.

```{r}
survkit_methods()
```

## A first fit

A Weibull accelerated-failure-time model on the lung-cancer data:

```{r}
fit <- survkit(survival::Surv(time, status) ~ age + sex, survival::lung,
               method = "weibull")
fit
```

`survkit_tidy()` gives the coefficient table with ratios and confidence
intervals; `survkit_curve()` gives predicted survival on a time grid.

```{r}
survkit_tidy(fit)
survkit_curve(fit, times = c(100, 300, 500))
```

## Switching method, same interface

The Cox proportional-hazards fit reads the same and tidies the same:

```{r}
cox <- survkit(survival::Surv(time, status) ~ age + sex, survival::lung,
               method = "cox")
survkit_tidy(cox)
```

## The small-sample gate

Survival models identify from events, not rows. Peduzzi et al.'s
events-per-variable (EPV) heuristic --
$$\mathrm{EPV} = \dfrac{\text{number of events}}{\text{number of predictors}},$$
with $\mathrm{EPV} \geq 10$ as the conventional floor -- is the standard guard
against coefficients and standard errors that cannot be trusted from too few
failures. On a small subset the gate flags the fit as under-powered:

```{r}
small <- survival::lung[1:25, ]
survkit_power(survival::Surv(time, status) ~ age + sex + ph.ecog, small)
```

`survkit()` runs this automatically and warns, so you can simplify, penalise, or
abstain rather than trust over-fit estimates.

## Visualising a fit

The step curve is the picture the summary numbers are drawn from. Fitting a
Kaplan-Meier baseline and plotting it against age group shows why: the two
groups' curves separate early and stay separated, which is the pattern a
Cox or Weibull regression on `age` is trying to summarise as one hazard ratio
or time ratio.

```{r fig-km, fig.cap = "Kaplan-Meier survival by age group (lung cancer trial). The confidence bands are pointwise 95 percent intervals. The older group's curve sits below the younger group's for the whole follow-up, so a single time-invariant hazard or time ratio is a defensible summary here -- the curves neither cross nor converge."}
lung2 <- survival::lung
lung2$age_group <- factor(ifelse(lung2$age >= stats::median(lung2$age),
                                 "older", "younger"),
                          levels = c("younger", "older"))
km_fit <- survival::survfit(survival::Surv(time, status) ~ age_group, lung2)
plot(km_fit, col = c("#1b9e77", "#d95f02"), lwd = 2, conf.int = TRUE,
    xlab = "Days", ylab = "Survival probability",
    main = "Kaplan-Meier survival by age group")
legend("bottomleft", legend = levels(lung2$age_group),
      col = c("#1b9e77", "#d95f02"), lwd = 2, bty = "n")
```

The two curves separate within the first 100 days and the gap holds out to
the point where the risk set is too thin to say more -- there is no crossing
that would signal a genuinely time-varying age effect.

## Competing risks

With a multi-state response (the event is a factor whose first level is
censoring), the Fine-Gray subdistribution hazard is one call:

```{r}
mg <- survival::mgus2
mg$etime <- with(mg, ifelse(pstat == 1, ptime, futime))
mg$event <- with(mg, factor(ifelse(pstat == 1, "pcm",
                            ifelse(death == 1, "death", "censor")),
                            levels = c("censor", "pcm", "death")))
fg <- survkit(survival::Surv(etime, event) ~ age + sex, mg,
              method = "fine_gray", cause = "pcm")
survkit_tidy(fg)
```

## Cumulative incidence

For competing risks the Aalen-Johansen cumulative incidence of each event is one
call, and it accounts for the competing event rather than treating it as
censoring:

```{r}
ci <- survkit_cuminc(survival::Surv(etime, event) ~ 1, mg)
head(ci)
```

Stacking the two causes' cumulative incidence shows how much of the total
event probability each accounts for at any time, and confirms that the two
curves sum to a proper (non-decreasing, bounded) overall failure probability:

```{r fig-cif, fig.cap = "Aalen-Johansen cumulative incidence of plasma-cell malignancy and death, stacked so the top line is the overall failure probability. Death dominates the overall event probability throughout follow-up, and pcm's cumulative incidence rises far more slowly -- exactly the separation the Fine-Gray hazard above is summarising."}
pcm <- ci[ci$cause == "pcm", c("time", "cif")]
death <- ci[ci$cause == "death", c("time", "cif")]
grid_t <- sort(unique(c(pcm$time, death$time)))
pcm_y <- stats::approx(pcm$time, pcm$cif, xout = grid_t, method = "constant",
                       rule = 2)$y
death_y <- stats::approx(death$time, death$cif, xout = grid_t,
                         method = "constant", rule = 2)$y
plot(grid_t, pcm_y + death_y, type = "s", col = "#7570b3", lwd = 2,
    xlab = "Days", ylab = "Cumulative incidence",
    main = "Stacked cumulative incidence: pcm and death",
    ylim = c(0, max(pcm_y + death_y)))
lines(grid_t, pcm_y, type = "s", col = "#1b9e77", lwd = 2)
legend("topleft", legend = c("pcm", "pcm + death"),
      col = c("#1b9e77", "#7570b3"), lwd = 2, bty = "n")
```

## Decision-relevant summaries

The restricted mean survival time (RMST) to a horizon $\tau$ is the area under
the survival curve on $[0, \tau]$,
$$\mathrm{RMST}(\tau) = \int_0^{\tau} S(t)\,dt,$$
a summary that stays defined when the median is never reached:

```{r}
km <- survkit(survival::Surv(time, status) ~ 1, survival::lung, method = "km")
survkit_rmst(km, tau = 365)
```

Predictions can be scored for discrimination and calibration. The concordance
index, the censoring-weighted (IPCW) Brier score, and its index of prediction
accuracy are returned together. The Brier score at horizon $t$ weights each
subject by the inverse probability of *not* being censored by $t$,
$$
BS(t) = \frac{1}{n}\sum_{i=1}^n \left[
  \frac{\big(0-\hat S(t\mid x_i)\big)^2}{\hat G(T_i)}\,
  \mathbf{1}(T_i \leq t,\ \delta_i = 1) +
  \frac{\big(1-\hat S(t\mid x_i)\big)^2}{\hat G(t)}\,
  \mathbf{1}(T_i > t) \right],
$$
where $\hat S$ is the predicted survival probability and $\hat G$ is the
Kaplan-Meier estimator of the censoring distribution (Graf et al. 1999); the
index of prediction accuracy, $IPA(t) = 1 - BS(t)/BS_0(t)$, rescales against a
covariate-free reference $BS_0$:

```{r}
survkit_score(cox, survival::lung, times = c(200, 400, 600))
```

## Triangulating across methods

Fitting the same question several ways is the honesty check survkit is built
for: methods that make different distributional assumptions about the same
hazard should still land near the same restricted mean survival time. When they
disagree by more than a `threshold` relative spread, the fit is flagged --
either because the shape of the hazard has been misspecified by at least one
of the methods, or because the compared quantities are not actually the same
estimand (a marginal curve compared against a curve at a reference covariate
profile, for instance). A flag is a prompt to check which of the two it is, not
a verdict on its own:

```{r}
tri <- survkit_triangulate(survival::Surv(time, status) ~ age + sex,
                           survival::lung,
                           methods = c("weibull", "cox", "km"), tau = 500)
tri$rmst
tri$flagged
```

Plotting each method's restricted mean as a point, with the disagreement band
drawn at `threshold` around their mean, makes the same table easier to read at
a glance -- this "triangulation fan" is the single picture that would have
shown a real disagreement immediately, rather than requiring a table lookup:

```{r fig-fan, fig.cap = "Triangulation fan: restricted mean survival time to tau = 500 days under three methods, with the shaded band spanning the mean plus or minus half the relative-spread threshold. All three estimates fall inside the band, matching the printed `flagged = FALSE` verdict above -- weibull, Cox and Kaplan-Meier concur within tolerance on this fit."}
band_half <- tri$threshold * mean(tri$rmst$rmst) / 2
band_mid <- mean(tri$rmst$rmst)
plot(seq_len(nrow(tri$rmst)), tri$rmst$rmst, pch = 19, cex = 1.4,
    col = "#1b9e77", xaxt = "n", xlab = "Method", ylab = "RMST (days)",
    xlim = c(0.5, nrow(tri$rmst) + 0.5),
    ylim = range(c(tri$rmst$rmst, band_mid - band_half, band_mid + band_half)),
    main = sprintf("Triangulated RMST at tau = %d", tri$tau))
axis(1, at = seq_len(nrow(tri$rmst)), labels = tri$rmst$method)
rect(0.5, band_mid - band_half, nrow(tri$rmst) + 0.5, band_mid + band_half,
    col = grDevices::adjustcolor("#7570b3", alpha.f = 0.15), border = NA)
abline(h = band_mid, lty = 2, col = "#7570b3")
```

## Extending the toolkit

Adding a method is one registry call; the public interface does not change.

```{r}
survkit_register("exp_survreg", kind = "parametric", backend = "survival",
  fit = function(formula, data, ...) {
    raw <- survival::survreg(formula, data = data, dist = "exponential", ...)
    list(raw = raw, coef = coef(raw), vcov = vcov(raw),
         loglik = as.numeric(logLik(raw)), n = nrow(data),
         n_events = NA_integer_, kind = "parametric", backend = "survival")
  })
"exp_survreg" %in% survkit_methods()$method
```
